The fraction of polar aligned circumbinary disks

### 2025-01-14 src/tex/ms.tex
Use 'narrow' instead of 'thin' to describe the ring

Thus, when the disk lifetime is longer than 
the alignment timescale,
%the viscous timescale \citep[which itself is similar to the precession timescale, see][]{bate2000}, 
it is appropriate to map the secular dynamics (which we call the dynamical state, e.g. nodal precession, libration or crescent orbits) to an end state of the disk (coplanar or polar, respectively). More challenging is mapping the parameters of the disk (e.g., binary eccentricity, initial inclination, disk mass, etc.) to a dynamical state. In general, this can be done using expensive 3D disk simulations (e.g., those shown in \citealt{martin2019}, \citealt{smallwood2019}, \& \citealt{abod2022}). It is not feasible to perform such simulations over a large parameter space, so we instead approximate a disk as a geometrically thin ring with angular momentum equivalent to that of the extended disk it represents. The literature provides a robust theoretical framework for the dynamics of such a ring, with analytical solutions \citep{farago2010,zanazzi2018,martin2019} as well as numerical $n$-body simulations \citep{abod2022,lepp2022} used to translate initial conditions into the ring's dynamical state.

In this study we use three methods, varying from purely analytic to $n$-body and statistical, to calculate $f_\text{polar}$ for a given population. In Section \ref{sec:analytic} we extend the analytic work to massive rings. Section \ref{sec:rk} 
describes the numerical integration of a set of coupled differential equations that describe the evolution of the binary-ring system in the secular approximation of the binary potential. In Section \ref{sec:reb} we treat the binary potential fully using the $n$-body code {\sc REBOUND} \citep{rebound}, finding only a negligible difference between this method and the much-faster secular approximation. We discuss our results in Section \ref{sec:results}, specifically looking at the effect that the binary eccentricity, relative angular momentum, and initial mutual inclination distributions have on the value of $f_\text{polar}$. Section \ref{sec:discussion} examines the consequences of our thin-ring approximation, and discusses what considerations must be made to apply our results to an extended disk. Finally, Section \ref{sec:conclusions} describes our overall conclusions. Code implementing the methods described in this paper is publicly available\footnote{\url{\codeurl}, or via \hfill \break \texttt{ pip install git+\codeurl} } as an installable Python package, and the scripts used to generate the figures along with the \LaTeX \space source are available\footnote{\ghurl} as well.

@@

To determine the polar fraction of a given population of geometrically thin rings we must consider three distributions: $p_j(j)$, $p_{e_\text{b}}(e_\text{b})$, and $p_i(i)$. If these distributions were known, it would be easy to calculate $f_\text{polar}$ for a population. However, none are well constrained, so in this section we calculate $f_\text{polar}$ considering one distribution at a time, drawing on parameterizations from relevant literature when available.

@@

If we consider a $p=1$ disk orbiting an equal mass binary with $e_\text{b}=0.4$ and $r_\text{in}/a_\text{b}=2.5$, then for a relatively thin ring ($r_\text{out}/r_\text{in} = 2$), we calculate $j_\text{max} = 0.17$. If the disk is allowed to be more extended (in this case $r_\text{out}/r_\text{in} = 10$), then we find $j_\text{max} = 2.8$.

@@

Note that equation (\ref{eq:jmax-steep}) is less than equation ({\ref{eq:jmax-shallow}}) by a factor of $(r_\text{out}/r_\text{in})^{2-p}$. If we consider the same example as in Section \ref{subsubsec:shallow-p}, except with $p=3$, we find that in the thin case ($r_\text{out}/r_\text{in} = 2$), $j_\text{max} = 0.040$, and is even smaller for $r_\text{out}/r_\text{in} = 10$, at  $j_\text{max} = 0.019$.

@@

In this study we attempt to quantify the fraction of polar disks in circumbinary systems using the dynamics of a thin ring. We do this analytically, numerically in the secular approximation, and numerically using the full binary potential. We find that the three methods produce values of the polar fraction that are identical within their uncertainties.

@@

We also discuss the limitations of our methods. First, we look at the effect of finite disk lifetime on the ability for a disk to complete its evolution to a stationary state. Disks with long evolutionary timescales remain with orbital parameters near their initial conditions for their entire lifetimes. We also look at limitations imposed by the fact that real disks are not geometrically thin rings, but have some radial extent. Real disks are likely to break if their radial extent is too large, and will become gravitationally unstable if all of their mass in concentrated in too thin of a ring. These two effects impose constraints on the relationship between $r_\text{in}$, $r_\text{out}$, and $j$.

>>>>

Thus, when the disk lifetime is longer than 
the alignment timescale,
%the viscous timescale \citep[which itself is similar to the precession timescale, see][]{bate2000}, 
it is appropriate to map the secular dynamics (which we call the dynamical state, e.g. nodal precession, libration or crescent orbits) to an end state of the disk (coplanar or polar, respectively). More challenging is mapping the parameters of the disk (e.g., binary eccentricity, initial inclination, disk mass, etc.) to a dynamical state. In general, this can be done using expensive 3D disk simulations (e.g., those shown in \citealt{martin2019}, \citealt{smallwood2019}, \& \citealt{abod2022}). It is not feasible to perform such simulations over a large parameter space, so we instead approximate a disk as a geometrically narrow ring with angular momentum equivalent to that of the extended disk it represents. The literature provides a robust theoretical framework for the dynamics of such a ring, with analytical solutions \citep{farago2010,zanazzi2018,martin2019} as well as numerical $n$-body simulations \citep{abod2022,lepp2022} used to translate initial conditions into the ring's dynamical state.

In this study we use three methods, varying from purely analytic to $n$-body and statistical, to calculate $f_\text{polar}$ for a given population. In Section \ref{sec:analytic} we extend the analytic work to massive rings. Section \ref{sec:rk} 
describes the numerical integration of a set of coupled differential equations that describe the evolution of the binary-ring system in the secular approximation of the binary potential. In Section \ref{sec:reb} we treat the binary potential fully using the $n$-body code {\sc REBOUND} \citep{rebound}, finding only a negligible difference between this method and the much-faster secular approximation. We discuss our results in Section \ref{sec:results}, specifically looking at the effect that the binary eccentricity, relative angular momentum, and initial mutual inclination distributions have on the value of $f_\text{polar}$. Section \ref{sec:discussion} examines the consequences of our narrow-ring approximation, and discusses what considerations must be made to apply our results to an extended disk. Finally, Section \ref{sec:conclusions} describes our overall conclusions. Code implementing the methods described in this paper is publicly available\footnote{\url{\codeurl}, or via \hfill \break \texttt{ pip install git+\codeurl} } as an installable Python package, and the scripts used to generate the figures along with the \LaTeX \space source are available\footnote{\ghurl} as well.

@@

To determine the polar fraction of a given population of geometrically narrow rings we must consider three distributions: $p_j(j)$, $p_{e_\text{b}}(e_\text{b})$, and $p_i(i)$. If these distributions were known, it would be easy to calculate $f_\text{polar}$ for a population. However, none are well constrained, so in this section we calculate $f_\text{polar}$ considering one distribution at a time, drawing on parameterizations from relevant literature when available.

@@

If we consider a $p=1$ disk orbiting an equal mass binary with $e_\text{b}=0.4$ and $r_\text{in}/a_\text{b}=2.5$, then for a relatively narrow ring ($r_\text{out}/r_\text{in} = 2$), we calculate $j_\text{max} = 0.17$. If the disk is allowed to be more extended (in this case $r_\text{out}/r_\text{in} = 10$), then we find $j_\text{max} = 2.8$.

@@

Note that equation (\ref{eq:jmax-steep}) is less than equation ({\ref{eq:jmax-shallow}}) by a factor of $(r_\text{out}/r_\text{in})^{2-p}$. If we consider the same example as in Section \ref{subsubsec:shallow-p}, except with $p=3$, we find that in the narrow case ($r_\text{out}/r_\text{in} = 2$), $j_\text{max} = 0.040$, and is even smaller for $r_\text{out}/r_\text{in} = 10$, at  $j_\text{max} = 0.019$.

@@

In this study we attempt to quantify the fraction of polar disks in circumbinary systems using the dynamics of a narrow ring. We do this analytically, numerically in the secular approximation, and numerically using the full binary potential. We find that the three methods produce values of the polar fraction that are identical within their uncertainties.

@@

We also discuss the limitations of our methods. First, we look at the effect of finite disk lifetime on the ability for a disk to complete its evolution to a stationary state. Disks with long evolutionary timescales remain with orbital parameters near their initial conditions for their entire lifetimes. We also look at limitations imposed by the fact that real disks are not geometrically narrow rings, but have some radial extent. Real disks are likely to break if their radial extent is too large, and will become gravitationally unstable if all of their mass in concentrated in too narrow of a ring. These two effects impose constraints on the relationship between $r_\text{in}$, $r_\text{out}$, and $j$.
